---
title: "pipeML"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pipeML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`pipeML` is a leakage-aware machine learning framework for classification and survival tasks on high-dimensional biological data, as described in [Hurtado and Pancaldi, 2026](https://doi.org/10.64898/2026.03.12.711429). Its central feature is custom cross-validation fold construction: features that depend on several samples at once (clusters, co-expression modules, enrichment scores, network features) are recomputed inside each fold, so the test samples never influence the features the model is trained on.

## Installation

To avoid GitHub API rate limit issues, set up a Personal Access Token (PAT) before installing:

```{r, eval=FALSE}
# install.packages(c("usethis", "gitcreds"))
usethis::create_github_token()
gitcreds::gitcreds_set()
```

Install `pipeML` from GitHub:

```{r, eval=FALSE}
# install.packages("pak")
pak::pkg_install("VeraPancaldiLab/pipeML")
```

Survival models also need the `censored` package (`install.packages("censored")`). You don't need to load it: `pipeML` loads it when a survival task runs.

## Core functions

- `compute_features.training.ML()`: train and tune models on a training set with repeated stratified k-fold cross-validation, and select the best one.
- `compute_prediction()`: evaluate the selected model on a test set (AUROC and AUPRC with bootstrap confidence intervals, or C-index).
- `compute_shap_values()`: explain the selected model with SHAP values.
- `compute_features.ML()`: training and prediction in one step, when the test set is already prepared.
- `get_curves()` and `plot_survival_performance()`: ROC and precision-recall curves, and Kaplan-Meier curves by predicted risk group.

Results (plots and the fold files of custom workflows) are written to a `Results/` folder in the working directory.

## Which workflow should I use?

| Workflow | When to use it | Tutorial |
|---|---|---|
| **Standard** | Your features are fixed values per sample (e.g. clinical variables, cell-type proportions computed per sample). | Classification, Survival analysis |
| **Leave-One-Dataset-Out** | Your samples come from several cohorts and you want to test generalization across them. | Multi-cohort validation (LODO) |
| **Custom folds** | Your features are computed from several samples at once (e.g. clustering, PCA, co-expression modules), so they must be recomputed inside each fold to avoid leakage. The feature construction can also have parameters to tune. | Leakage-aware custom cross-validation |

All workflows work for classification and survival tasks, and all of them support prediction on a test set and SHAP values.

## Tutorials

Step-by-step tutorials are available in the **Articles** section of the navigation bar:

- **Classification** — train, tune and select classification models, predict on a test set, and train and predict in one step
- **Survival analysis** — the same workflow for time-to-event outcomes, with Kaplan-Meier curves by predicted risk group
- **Interpreting models with SHAP values** — explain the selected model globally and for single samples
- **Multi-cohort validation (LODO)** — leave-one-dataset-out validation across cohorts
- **Leakage-aware custom cross-validation** — recompute sample-dependent features inside each fold, with or without tunable parameters

## Citation

If you use `pipeML` in a scientific publication, please cite:

> Hurtado, M., & Pancaldi, V. (2026). *A new pipeline for cross-validation fold-aware machine learning prediction of clinical outcomes addresses hidden data-leakage in omics based 'predictors'.* bioRxiv. <https://doi.org/10.64898/2026.03.12.711429>

`pipeML` is built on top of `caret`, `tidymodels`, `parsnip` and `censored`; please also cite the packages of the algorithms you use.
